---
title: "Statistical Inference Project Part 2"
author: "Devon Terrell"
date: "Thursday, December 18, 2014"
output: html_document
---

## Exploring the ToothGrowth Dataset

The ToothGrowth dataset consists of measurements of the lengths of teeth in 10 guinea pigs which were given varying doses of Vitamin C (0.5, 1.0, and 2.0 mg) using two different delivery mechanisms (orange juice and asorbic acid).

```{r}
tg <- ToothGrowth
combined_mean <- mean(tg$len)
combined_var <- var(tg$len)
hist(tg$len, main = "Histogram of Tooth Length")
abline(v = combined_mean, col = "red", lwd = 2)
```

Statistic | Value
--------- | ------
Mean      | `r combined_mean`
Variance  | `r combined_var`

## Separating Out Observation Classes

The data consists of six classes of observations, one for each combination of dose and supplement type.  Each class has its own histogram of observed lengths and assoicated descriptive statistics.

```{r}
parted_data = split(tg, list(tg$supp, tg$dose))
par(mfrow=c(2,3))
silence <- mapply(function(c,i){hist(c$len, main = paste("Histogram of", i))}, parted_data, names(parted_data))
```

```{r}
aggregate(tg$len, by=list(tg$supp, tg$dose), function(x){c(MU=mean(x), SD=sd(x))})
```